---
title: "physical dataset preparation"
author: "Owen Liu"
date: "September 1, 2016"
output: html_document
---

## Preparation of Physical Datasets

The following code takes raw physical datasets and matches them to the same timescale as the SNI community monitoring datasets.  There are four physical variables to be matched:

* The Multivariate ENSO index
* The Pacific Decadal Oscillation
* The North Pacific Gyre Oscillation
* Sea Surface Temperature and Significant Wave Height

For each, we match them to the "Periods" from the SNI benthic monitoring datasets, which are 6-month intervals between 1980 and 2011.

```{r setup, include=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)
library(tseries)
library(rEDM)
library(reshape2)

#### Data ####
# The raw data has been processed into a set of normalized time series elsewhere.
W_D <- getwd()

# Multivariate ENSO index (MEI)
mei <- read.csv(paste0(W_D,"/data/mei.csv"))

# Pacific Decadal Oscillation
pdo <- read.csv(paste0(W_D,"/data/PDO.csv"))

# North Pacific Gyre Oscillation
npgo <- read.csv(paste0(W_D,"/data/npgo.csv"))

# SST and wave height, from CDIP
waveSST <- read.csv(paste0(W_D,"/data/Begg_SNI_CDIP.csv"))
```

### Create a Period to Year/Month matching key

In order to line up the data from all of these sources, we need a key that matches year/month to periods. Very generally, SNI data have been collected every 6 months from Fall 1980 (Period 1) to Fall 2011 (Period 63).  For the purposes of matching, we'll include averaged data from the previous 4 months to apply to each period.  That means that the Spring periods will reflect Dec-Mar averages of the physical variables, and Fall will represent Jun-Sep values.

```{r period date match}
# Have to line up the periods such that December is included in the next Spring's period
period.key <- data_frame(year=rep(1980:2011,each=12),month=rep(1:12,32),period=c(rep(NA,5),rep(1,5),rep(2:63,each=6),rep(64,2)))

# Set months that are not in Dec-Mar or Jun-Sep as NA so their data aren't included
period.key$period[!period.key$month %in% c(12,1,2,3,6,7,8,9)] <- NA
```

### Reorganize the Data

Now we can add a period identifier to the physical data. One dataset at a time

```{r period match mei}
# MEI data
glimpse(mei)

# These data are for every year since 1950, and bimonthly MEI index values. We'll set the months to the normal integers, keeping in mind that they represent a sliding scale (http://www.esrl.noaa.gov/psd/enso/mei/)

# Convert from wide to long form and add integer month identifier
mei.long <- mei %>% 
  gather(key=month.txt,value=mei,DecJan:NovDec) %>%
  arrange(Year) %>%
  mutate(month.num= rep(1:12,67))

# match to period key and remove irrelevant data
mei.per <- mei.long %>% 
  left_join(period.key,by=c("Year"="year","month.num"="month")) %>%
  filter(!is.na(period)) %>%
  
  # for each period, average the MEI value to make just one value for each period
  group_by(period) %>%
  summarise(mei.mean=mean(mei))

## Plot of output
ggplot(mei.per,aes(x=period,y=mei.mean)) +
  geom_line()+
  xlab("Period")+
  ylab("Multivariate ENSO Index")+
  ggtitle("MEI over SNI monitoring periods")+
  geom_hline(yintercept=0,linetype=2)
```

Next is the Pacific Decadal Oscillation data

```{r pdo match}
glimpse(pdo)

# Similar procedure-- convert to long form add a month identifier, and then match to periods
pdo.long <- pdo %>%
  gather(key=month.txt,value=pdo,JAN:DEC) %>%
  arrange(YEAR) %>%
  mutate(month.num=rep(1:12,117))

# Match to periods and filter
pdo.per <- pdo.long %>% 
  left_join(period.key,by=c("YEAR"="year","month.num"="month")) %>%
  filter(!is.na(period)) %>%
  
  # for each period, average the PDO value to make just one value for each period
  group_by(period) %>%
  summarise(pdo.mean=mean(pdo))

## Plot of output
ggplot(pdo.per,aes(x=period,y=pdo.mean)) +
  geom_line()+
  xlab("Period")+
  ylab("PDO Index")+
  ggtitle("PDO over SNI monitoring periods")+
  geom_hline(yintercept=0,linetype=2)
```

Now, the North Pacific Gyre Oscillation (NPGO)

```{r npgo match}
glimpse(npgo)

# NPGO is already in long form, so we don't need to convert it

# Match to periods and filter
npgo.per <- npgo %>% 
  left_join(period.key,by=c("YEAR"="year","MONTH"="month")) %>%
  filter(!is.na(period)) %>%
  
  # for each period, average the NPGO value to make just one value for each period
  group_by(period) %>%
  summarise(npgo.mean=mean(NPGO.index))

npgo.yr <- npgo %>% filter(YEAR>1979) %>% unite(date,MONTH,YEAR,sep=" ")

npgo.yr$date <- as.yearmon(npgo.yr$date,format="%m %Y") %>% as.Date()

ggplot(npgo.yr,aes(x=date,y=NPGO.index)) +
  geom_line(col="darkgreen",size=1.5)+
  scale_x_date()+
  xlab("Year")+
  ylab("NPGO Index")+
  geom_hline(yintercept=0,linetype=2)+
  theme_minimal()

## Plot of output
ggplot(npgo.per,aes(x=period,y=npgo.mean)) +
  geom_line(col="darkgreen",size=2)+
  xlab("Period")+
  ylab("NPGO Index")+
  geom_hline(yintercept=0,linetype=2)
```

Finally, SST and significant wave height.  There are some missing data here, but we'll just include NAs for the periods without data.  For signficant wave height, we use the maximum from the previous four months for each period, as that seems to have a more predicatable effect than average significant wave height.

```{r SST and waves match}
glimpse(waveSST)

# separate into two separate variables. Hs denotes significant, wave height:
# Significant wave height in meters; derived from the zeroth moment of the reported energy spectrum. 
# Described as the "average height of the one third highest waves in the record"
waves <- waveSST %>% select(Dataset,Year,Month,Mean_Hs)

# match to periods
waves.per <- waves %>%
  left_join(period.key,by=c("Year"="year","Month"="month")) %>%
  filter(!is.na(period)) %>%
  
  # group by period, and MAX
  group_by(period) %>%
  summarise(waves.max=max(Mean_Hs))

# plot of wave height by period
ggplot(waves.per,aes(x=period,y=waves.max)) +
  geom_line()+
  xlab("Period")+
  ylab("Maximum Significant Wave Height")+
  ggtitle("Max Wave Height over SNI monitoring periods")

# SST
sst <-  waveSST %>% select(Dataset,Year,Month,Mean_SST)

# match to periods
sst.per <- sst %>%
  left_join(period.key,by=c("Year"="year","Month"="month")) %>%
  filter(!is.na(period)) %>%
  
  # group by period, and MAX
  group_by(period) %>%
  summarise(sst.mean=mean(Mean_SST))

waves.yr <- waves %>% filter(Year>1979) %>% unite(date,Month,Year,sep=" ")

waves.yr$date <- as.yearmon(waves.yr$date,format="%m %Y") %>% as.Date()

ggplot(waves.yr,aes(x=date,y=Mean_Hs)) +
  geom_line(col="blue",size=1)+
  scale_x_date()+
  xlab("Year")+
  ylab("Mean Wave Height")+
  theme_minimal()
```

### Data join and export
With the common period identifier, we can now join and export all the data

```{r data join and export}
phys.dat <- full_join(mei.per,pdo.per) %>% full_join(npgo.per) %>% 
  full_join(waves.per) %>% full_join(sst.per) %>%
  filter(period != 64)

# We can also normalize the variables to match dimensionally with our biological data
normalise <- function(vec) (vec-mean(vec,na.rm=T))/sd(vec,na.rm=T)

phys.dat.norm <- phys.dat %>% mutate_each(funs(normalise),-period)

## Plot of all normalized physical variables
phys.dat.norm.long <- gather(phys.dat.norm,key=var,value=val,mei.mean:sst.mean)
ggplot(phys.dat.norm.long,aes(x=period,y=val))+
  geom_line(aes(col=var))+
  ggtitle("SNI Physical Variables by Monitoring Period")+
  xlab("Period")+
  ylab("Normalized Value")+
  geom_hline(yintercept=0,linetype=2)

# export data
write.csv(phys.dat,file=paste0(W_D,"/data/phys_vars_combined.csv"),row.names = F)
write.csv(phys.dat.norm,file=paste0(W_D,"/data/phys_vars_combined_norm.csv"),row.names = F)

```